---
title: 'Project 1: Exploratory Data Analysis'
author: "Andre Ferreira AF32263"
date: '4/2/21'
output:
  pdf_document:
    toc: no
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))

```

## Data Wrangling and Data Exploration

### Instructions
A knitted R Markdown document (ideally HTML) and the raw R Markdown file (as .Rmd) should both be submitted to Canvas by 11:59pm on the due date. These two documents will be graded jointly, so they must be consistent (i.e., don’t change the R Markdown file without also updating the knitted document).

The text of the document should provide a narrative structure around your code/output. All results presented must have corresponding code. Any answers/results/plots etc. given without the corresponding R code that generated the result will not be considered. Furthermore, all code contained in your final project document must work correctly (knit early, knit often)! Please do not include any extraneous code or code which produces error messages. (Code that produces warnings is acceptable, as long as you understand what the warnings mean!)

### Find data:

Find two (!) datasets with one variable in common (e.g., dates, times, states, counties, countries, sports players), both with at least 50 observations (i.e., rows) in each. Please think very carefully about whether it makes sense to combine your datasets! If you find one dataset with 50 patients and it has their age, and you find another dataset with 50 *different* patients that has their ages, it makes no sense to join them based on age (you would just be pairing up random people of the same age).

When combined, the resulting/final dataset must have **at least 4 different variables (at least 3 numeric) in addition to the common variable** (i.e., five variables total).

You can have as many variables as you would like! If you found two datasets that you like but they don't have enough variables, find a third dataset with the same common variable and join all three.

```{r}
```
*For this project, I will be using two open datasets that I found from Data.World. The first dataset is a subset from a larger dataset titled "Crime in US 2016 Offenses", a dataset that provides information on the quantity and types of crime in the United States in 2016, categorized by State. The specific subset that I chose for this report lists the population of each state, the total number of aggravated assaults by state, and the number of aggravated assaults performed with a specific weapon (firearms, knife, etc.). The second dataset that I chose for this report contains information on the marijuana legalization status of every state, which places each state in one of three categories: illegal, legalized for recreational use, legalized for medical use.*

*In this project, I will be merging the two datasets through "States", and I am interested in observing (if any) correlation between legalization status and the number of aggravated assaults in a year while controlling for population.*




### Guidelines

1. If the datasets are not tidy, you will need to reshape them so that every observation has its own row and every variable its own column. If the datasets are both already tidy, you will make them untidy with `pivot_wider()/spread()` and then tidy them again with `pivot_longer/gather()` to demonstrate your use of the functions. It's fine to wait until you have your descriptives to use these functions (e.g., you might want to pivot_wider() to rearrange the data to make your descriptive statistics easier to look at); it's fine long as you use them at least once!

    
```{r}

#The datasets are already tidy, so in this code I am converting the rows into columns (pivot-wider), and then converting these columns back to rows (pivot_longer) to demonstrate use of the tidying functions. I also removed NAs after pivoting longer, as each state was repeated multiple times when attempting to format.
library(tidyverse)
mjlaws <- read.csv("~/Lab 7/state_marijuana_laws_10_2016.csv")
assault <- read.csv("~/Lab 7/assault1.csv")

assault <- assault[-c(51,52),]
assault1 <- assault %>% pivot_wider(names_from = State, values_from=Firearms)
assault1 <- assault1 %>% pivot_longer(c("Alabama":"Wyoming"), names_to="States", values_to="Firearms")
assault1 <- assault1 %>% na.omit()


mjlaw1 <- mjlaws %>% pivot_wider(names_from = State, values_from= Legalization.Status)
mjlaw1 <- mjlaw1 %>% pivot_longer(c("Connecticut":"Washington"), names_to="States", values_to="Legalization.Status")
mjlaw1 <- mjlaw1 %>% na.omit()

```
    
    

2. Join your 2+ separate data sources into a single dataset based on a common ID variable! If you can't find a good pair datasets to join, you may split one main dataset into two different datasets with a common ID variable in each, and then join them back together based on that common ID, but this is obviously less than ideal.

```{r}
fulldat <- full_join(assault, mjlaws, by="State")
fulldat1 <- fulldat %>% select(-c("Knives.or..cutting..instruments", "Other..weapons", "Personal..weapons", "Agency..count"))
fulldat <-fulldat1
```

*I performed a full_join to merge both datasets by State to organize the data better. In the "Mjlaw" dataset, there was 1 observation, not including each state. In the "assault" dataset, there were 7 observations, not including each state. I also decided to remove 4 observations from the merged dataset as seen above, as I did not find them important nor interesting.*

```{r}
#Converting numbers in data to numerical values and removing comma (the numbers were classified as character values).
replaceCommas<-function(x){
  x<-as.numeric(gsub("\\,", "", x))
}
fulldat$Total..aggravated..assaults1 <- replaceCommas(fulldat$Total..aggravated..assaults1)

fulldat$Firearms <- replaceCommas(fulldat$Firearms)
fulldat$Population <- replaceCommas(fulldat$Population)
```

3. Create summary statistics

    
```{r}
library(dplyr)
#Arrange
fulldat %>% arrange(Total..aggravated..assaults1)

#Pivot Wider
fulldat %>% pivot_wider(State, names_from=State, values_from=Total..aggravated..assaults1)

#Grouping
fulldat %>% group_by(Legalization.Status) %>% arrange(Total..aggravated..assaults1)

#Filter
fulldat %>% group_by(State) %>% filter(Legalization.Status=="illegal")

#Select
fulldat %>% select(Total..aggravated..assaults1, Legalization.Status) %>% arrange(desc(Total..aggravated..assaults1))

#Summarize
fulldat %>% group_by(Legalization.Status) %>% summarize(mean_assault=mean(Total..aggravated..assaults1))

#Mutate
fulldat <- fulldat %>% mutate(proportion.firearm=Firearms/Total..aggravated..assaults1)

fulldat <- fulldat %>% mutate(assault.percapita = Total..aggravated..assaults1/Population)




```

    
    - Create summary statistics (`mean, sd, var, n, quantile, min, max, n_distinct, cor`, etc) for each of your numeric variables both overall and after grouping by one of your categorical variables (either together or one-at-a-time; if you have two categorical variables, try to include at least one statistic based on a grouping of two categorical variables simultaneously). If you do not have any categorical variables, create one using mutate (e.g., with `case_when` or `ifelse`) to satisfy the `group_by` requirements above. Ideally, you will find a way to show these summary statistics in an easy-to-read table (e.g., by reshaping). (You might explore the kable package for making pretty tables!) If you have lots of numeric variables (e.g., 10+), or your categorical variables have too many categories, just pick a few (either numeric variables or categories of a categorical variable) and summarize based on those. It would be a good idea to show a correlation matrix for your numeric variables (you will need it to make one of your plots).
 
 
```{r}

# Summary Statistics Overall
fulldat %>% summarise(mean(Total..aggravated..assaults1), sd(Total..aggravated..assaults1), var(Total..aggravated..assaults1), n(), quantile(Total..aggravated..assaults1), n_distinct(Total..aggravated..assaults1), min(Total..aggravated..assaults1), max(Total..aggravated..assaults1))

fulldat %>% summarise(mean(Population), sd(Population), var(Population), n(), quantile(Population), n_distinct(Population), min(Population), max(Population))

fulldat %>% summarise(mean(Firearms), sd(Firearms), var(Firearms), n(), quantile(Firearms), n_distinct(Firearms), min(Firearms), max(Firearms))

fulldat %>% summarise(mean(assault.percapita), sd(assault.percapita), var(assault.percapita), n(), quantile(assault.percapita), n_distinct(assault.percapita), min(assault.percapita), max(assault.percapita))

#Correlation Matrix for Numerical Values
cordat <- fulldat %>% select_if(is.numeric) %>% cor(use="pair")

#Summary Statistics by Group
fulldat %>% group_by(Legalization.Status) %>% summarise(mean(Total..aggravated..assaults1), sd(Total..aggravated..assaults1), var(Total..aggravated..assaults1), n(), quantile(Total..aggravated..assaults1), n_distinct(Total..aggravated..assaults1), min(Total..aggravated..assaults1), max(Total..aggravated..assaults1))

fulldat %>% group_by(Legalization.Status) %>% summarise(mean(Population), sd(Population), var(Population), n(), quantile(Population), n_distinct(Population), min(Population), max(Population))

fulldat %>% group_by(Legalization.Status) %>% summarise(mean(Firearms), sd(Firearms), var(Firearms), n(), quantile(Firearms), n_distinct(Firearms), min(Firearms), max(Firearms))

fulldat %>% group_by(Legalization.Status) %>% summarise(mean(assault.percapita), sd(assault.percapita), var(assault.percapita), n(), quantile(assault.percapita), n_distinct(assault.percapita), min(assault.percapita), max(assault.percapita))


```
 *First, I calculated summary statistics for the overall data without grouping. The mean total number of aggravated assaults per state is 14615.76 with a standard deviation of 19,226.9. There is a wide range between states with the lowest state having 302 aggravated assaults and the highest state havingn 104,295 aggravated assaults. States also vary highly in population, with the average population of all 50 states being 5,844,783 people. Because of the large variation in state population, I used the 'mutate' function to generate a new numerical variable, "assaults.percapita", which represents tha number of aggravated assaults divided by the total population in that state.*
 
 *After calculating summary statistics for the overall data, I then calculated summary statistics for each numerical variable while categorizing by marijuana legalization status. After doing this, I found that states with marijuana legalized for recreational use show the highest number of assaults per capita. States where marijuana is illegal show the lowest number of assaults percapita.*
 
 
4. Make visualizations (three plots)

    
```{r}
#Tidying up Correlations
tidycor <- cordat %>% as.data.frame %>% rownames_to_column("var1") %>% pivot_longer(-1,names_to="var2", values_to="correlation")

#Creating Correlation Heatmap
tidycor %>% ggplot(aes(var1,var2,fill=correlation))+geom_tile()+scale_fill_gradient2(low="red", mid="white", high="green")+geom_text(aes(label=round(correlation,2)),color="black",size =3) + theme(axis.text.x = element_text(angle=45, vjust = 1, size=6.5, hjust=1)) + ggtitle("Correlation Heatmap")
```
*This heatmap shows the correlation between all numerical variables in the dataset. It appears that the number of aggravated assaults is highly correlated to the population of each state (Cor=0.97). It also appears that the number of aggravated assaults is highly correlated to the use of firearms during assaults (Cor=0.94). There is a weak correlation between certain variables shown in the heatmap, which is understandable considering 'assault.percapita' and 'proportion.firearm' are both functions of other variables. A correlation in these variables is not expected. *
```{r}
#Mean Assaults per capita, grouped by Marijuana Legalization Status
fulldat %>% ggplot(aes(x=Legalization.Status, y=assault.percapita))+geom_bar(stat="summary", fun=mean, aes(fill=Legalization.Status))+geom_errorbar(stat="summary", fun.data=mean_se) + theme(axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank()) +ggtitle("Mean Assaults Per Capita by legalization status")
```
*This chart maps the assaults per capita for states grouped by their marijuana legalization status. From intial appearance, it appears that states where marijuana is legalized for recreational use have a higher count of assaults per capita than states where marijuana is only legalized for medical use and states where marijuana is not legalized at all. After controlling for standard error, it appears that there is nno significant difference in the number of assaults per capita between states of different legalization statuses.*

```{r}
#Total Aggravated Assaults by Population, grouped by Marijuana Legalization Status
fulldat %>% ggplot(aes(x=Population/1000000, y=Total..aggravated..assaults1, color=Legalization.Status)) +geom_point(size=1) + scale_x_continuous(name="Population (million)") +scale_y_continuous(name="Total Aggravated Assaults", breaks=seq(0,150000,5000)) + ggtitle("Total Aggravated Assaults by population, grouped by MJ Legalization Status")
```
*This chart displays how the total number of aggravated assaults changes with respect to population in all 50 states. They are also color-coded by marijuana legalization status, which makes it easier to visualize how legalization status influences (or is influenced by) the total number of aggravated assaults as well as population. From the data, it can be seen that there is little correlation between legalization status and population, and little correlation between legalizationn status and the number of aggravated assaults.*
    
5. Perform k-means/PAM clustering or PCA on (at least) your numeric variables.

```{r}
#Clustering with Categorical data
library(cluster)
cat_dat<-fulldat%>%mutate_if(is.character,as.factor)%>%column_to_rownames("State")%>%select(c(1:4))

#computing gower dissimilarities and using silhouette width to pick number of clusters
gower1<-daisy(cat_dat,metric="gower")

sil_width<-vector()
for(i in 2:10){
pam_fit <- pam(gower1, diss = TRUE, k = i)
sil_width[i] <- pam_fit$silinfo$avg.width
}
ggplot()+geom_line(aes(x=1:10,y=sil_width))+scale_x_continuous(name="k",breaks=1:10)

```
*The data above shows that 4 clusters has the highest average silhouette width, meaning that 4 clusters is the ideal cluster count that I will be using for PAM clustering.*

```{r}
#Performing pam with ideal number of clusters from above. Telling pam to use dissimilarities
pam3 <- pam(gower1, k = 4, diss = T) 


#Fnding which State is most representative of each cluster
fulldat %>% slice(pam3$id.med)

#Determining the Characteristics of each cluster
cat_dat%>%mutate(cluster=factor(pam3$clustering))%>%select(cluster,Legalization.Status)%>%
group_by(cluster,Legalization.Status)%>%summarize(n=n())%>%mutate(prop=n/sum(n,na.rm=T))%>%
pivot_wider(-n,names_from=Legalization.Status,values_from=prop)


```
*From the data above, it appears that Oklahoma, Nevada, New Mexico, and Texas are the representatives for each cluster. Cluster 1 corresponds to states where marijuana legalization status is 'illegal'. Cluster 2 are states where marijuana is legal for recreational use. Cluster 3 is where marijuana is legal for medical use. Cluster 4 contains states from all three categories, but is mainly categorized by highly populated states. *

```{r}
#Computing Goodness of Fit using silhouette plot of each cluster.
plot(pam3,which=2)

```

*The plot above shows silhouettes with an average width of 0.71, meaning a strong structure has been found when categorizing each cluster. Cluster 4 appears to have a very weak structure. In a repeat experiment, 3 clusters would probably be more accurate for categorizing the data.*

```{r}
#Generating final visualization for pam clustering
library(GGally)
fulldat%>%mutate(cluster=as.factor(pam3$clustering))%>%
ggpairs(columns = 2:5,aes(color=cluster)) + theme(axis.text.x = element_text(angle=45, vjust = 1, size=6.5, hjust=1))
```
*In the data above, it can be seen that each cluster is accurately categorized by legalization status, with the exception of cluster 4 which contains States from all three categories. States in cluster 4 appear to be highly populated states which also have greater cases of aggravated assaults. States where marijuana is illegal (Cluster1) appear to have the weakest correlation between population and total aggravated assault, confirming the earlier finding when plotting assaults percapita by legalization status. However, it appears that in the states where marijuana is illegal, a greater proportion of aggravated assaults are performed using firearms than states in other categories. In contrast, highly populated states (Cluster 4) have the most aggravated assaults as expected, but a lesser proportion of those assaults are performed using firearms. As a final conclusion, it appears that aggravated assaults are highly correlated with population but are not significantly correlated with the legalization status of marijuana.*

<P style="page-break-before: always">
\newpage
    
### Rubric

Prerequisite: Finding appropriate data from at least two sources per the instructions above: Failure to do this will result in a 0! You will submit a .Rmd file and a knitted document (html/pdf).

#### 0. Introduction (5  pts)

- Write a narrative introductory paragraph or two describing the datasets you have chosen, the variables they contain, how they were acquired, and why they are interesting to you. Expand on potential associations you may expect, if any.

#### 1. Tidying: Rearranging Wide/Long (10 pts)

- Tidy the datasets (using the `tidyr` functions `pivot_longer`/`gather` and/or `pivot_wider`/`spread`) 
- If you data sets are already tidy, be sure to use those functions somewhere else in your project (e.g., for rearranging summary statistics)
- Document the process (describe in words what was done)
    
#### 2. Joining/Merging (10 pts)

- Join your datasets into one using a `dplyr` join function
- If you have multiple observations on the joining variable in either dataset, fix this by collapsing via summarize
- Discuss the process in words, including why you chose the join you did
- Discuss which cases were dropped, if any, and potential problems with this

#### 3. Wrangling (40 pts)

- Use all six core `dplyr` functions in the service of generating summary tables/statistics (12 pts)
    - Use mutate at least once to generate a variable that is a function of at least one other variable

- Compute summary statistics for each of your variables using `summarize` alone and with `group_by` (if you have more than 10 variables, fine to just focus on 10) (20 pts)
    - Use at least 5 unique functions inside of summarize (e.g., mean, sd)
    - For at least 2, use summarize after grouping by a categorical variable. Create one by dichotomizing a numeric if necessary
    - If applicable, at least 1 of these should group by two categorical variables

- Summarize the procedure and discuss all (or the most interesting) results in no more than two paragraphs (8 pts)


#### 4. Visualizing (30 pts)

- Create a correlation heatmap of your numeric variables the way we did in class

- Create two more effective, polished plots with ggplot

    - Each plot should map 3+ variables to aesthetics 
    - Each plot should have a title and clean labeling for all mappings
    - Change at least one default theme element and color for at least one mapping per plot
    - For at least one plot, add more tick marks (x, y, or both) than are given by default
    - For at least one plot, use the stat="summary" function
    - Supporting paragraph or two (for each plot) describing the relationships/trends that are apparent
    
#### 5. Dimensionality Reduction (30 pts) 

- Either k-means/PAM clustering or PCA (inclusive "or") should be performed on at least three of your variables (3 is just the minimum: using more/all of them will make this much more interesting!)

    - All relevant steps discussed in class (e.g., picking number of PCs/clusters)
    - A visualization of the clusters or the first few principal components (using ggplot2)
    - Supporting paragraph or two describing results found, interpreting the clusters/PCs in terms of the original variables and observations, discussing goodness of fit or variance explained, etc.


#### 6. Neatness, Holistic/Discretionary Points (5 pts)

- Keep things looking nice! Your project should not knit to more than 30 or so pages (probably closer to 10-20)! You will lose points if you print out your entire dataset(s), have terrible formatting, etc. If you start your project in a fresh .Rmd file, you are advised to copy set-up code below and include it: this will do things like automatically truncate if you accidentally print out a huge dataset, etc. Imagine this is a polished report you are giving to your PI or boss to summarize your work researching a topic.

```{r eval=F}
## paste this chunk into the ```{r setup} chunk at the top of your project 1 .Rmd file

knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```


### Where do I find data?

OK, brace yourself!

You can choose ANY datasets you want that meet the above criteria for variables and observations. I'm just sitting here but off the top of my head, if you are into amusement parks, you could look at amusement-park variables, including ticket sales per day etc.; then you could join this by date in weather data. If you are interested in Game of Thrones, you could look at how the frequency of mentions of character names (plus other character variables) and the frequency of baby names in the USA...You could even take your old Biostats data and merge in new data (e.g., based on a Google forms timestamp).

You could engage in some "me-search": You can [request your Spotify data](https://support.spotify.com/ca-en/article/data-rights-and-privacy-settings/) or [download Netflix viewing activity](https://help.netflix.com/en/node/101917), Amazon purchase history, etc. You can use your Google Fit/Fitbit/Apple watch data, etc. These can be combined (e.g., with each other, with other data sources).

You can make it as serious as you want, or not, but keep in mind that you will be incorporating this project into a portfolio webpage for your final in this course, so choose something that really reflects who you are, or something that you feel will advance you in the direction you hope to move career-wise, or something that you think is really neat. On the flip side, regardless of what you pick, you will be performing all the same tasks, so it doesn't end up being that big of a deal.

If you are totally clueless and have no direction at all, log into the server and type 

```{R eval=F}
data(package = .packages(all.available = TRUE))
```

This will print out a list of **ALL datasets in ALL packages** installed on the server (a ton)! Scroll until your eyes bleed! Actually, do not scroll that much... To start with something more manageable, just run the command on your own computer, or just run `data()` to bring up the datasets in your current environment. To read more about a dataset, do `?packagename::datasetname`. 

If it is easier for you, and in case you don't have many packages installed, a list of R datasets from a few common packages (also downloadable in CSV format) is given at the following website: https://vincentarelbundock.github.io/Rdatasets/datasets.html (including types/numbers of variables in each)

- A good package to download for fun/relevant data is `fivethiryeight`. Just run `install.packages("fivethirtyeight"), load the packages with `library(fivethirtyeight)`, run `data()`, and then scroll down to view the datasets. Here is an online list of all 127 datasets (with links to the 538 articles). Lots of sports, politics, current events, etc: https://cran.r-project.org/web/packages/fivethirtyeight/vignettes/fivethirtyeight.html

- If you have already started to specialize (e.g., ecology, epidemiology) you might look at discipline-specific R packages (vegan, epi, respectively). We will be using some tools from these packages later in the course, but they come with lots of data too, which you can explore according to the directions above

- However, you *emphatically DO NOT* have to use datasets available via R packages! In fact, I would much prefer it if you found the data from completely separate sources and brought them together (a much more realistic experience in the real world)! You can even reuse data from your SDS328M project, provided it shares a variable in common with other data which allows you to  merge the two together (e.g., if you still had the timestamp, you could look up the weather that day: https://www.wunderground.com/history/). If you work in a research lab or have access to old data, you could potentially merge it with new data from your lab!

- Here is a curated list of interesting datasets (read-only spreadsheet format): https://docs.google.com/spreadsheets/d/1wZhPLMCHKJvwOkP4juclhjFgqIY8fQFMemwKL2c64vk/edit

- Here is another great compilation of datasets: https://github.com/rfordatascience/tidytuesday

- Here is the UCI Machine Learning Repository: https://archive.ics.uci.edu/ml/index.php

    - See also https://en.wikipedia.org/wiki/List_of_datasets_for_machine-learning_research#Biological_data

- Here is another good general place to look: https://www.kaggle.com/datasets

- To help narrow your search down or to see interesting variable ideas, check out https://www.tylervigen.com/spurious-correlations. This is the spurious correlations website, and it is fun, but if you look at the bottom of each plot you will see sources for the data. This is a good place to find very general data (or at least get a sense of where you can scrape data together from)!

- If you are interested in medical data, check out www.countyhealthrankings.org

- If you are interested in scraping UT data, the university makes *loads* of data public (e.g., beyond just professor CVs and syllabi). Check out all the data that is available in the statistical handbooks: https://reports.utexas.edu/statistical-handbook

##### Broader data sources:

[Data.gov](www.data.gov) 186,000+ datasets!

[Social Explorer](Social Explorer) is a nice interface to Census and American Community Survey data (more user-friendly than the government sites). May need to sign up for a free trial.

[U.S. Bureau of Labor Statistics](www.bls.gov)

[U.S. Census Bureau](www.census.gov)

[Gapminder](www.gapminder.org/data), data about the world.

...





